\section{Convergence: Integrating External Agent Frameworks}\label{sec:convergence}

The preceding sections develop Operon's structural analysis, epistemic topology,
and formal verification foundations. This section summarizes how these tools
extend to external agent orchestration systems through a typed adapter
architecture. A standalone companion paper provides the full treatment, including
implementation details, all 98 examples, and complete TLA+ specification
listings; this section is self-contained but deliberately concise.

\subsection{Five-Layer Architecture}

The convergence architecture stacks five functional layers, each building on the
one below. At the base, Operon provides the structural layer: typed wiring
diagrams, epistemic theorems, and the resource-aware executor. Above it, three
orchestration runtimes---Swarms~\cite{swarms} for enterprise-scale multi-agent
workflows, DeerFlow~\cite{deerflow} for session-based research pipelines, and
Ralph~\cite{ralph} for hat-based event-driven agents---provide the operational
execution surface. The thinking layer (AsyncThink) adds fork/join concurrency
within individual stages, enabling speculative parallel reasoning without
altering the outer topology. AnimaWorks~\cite{animaworks} contributes the
cognitive layer: developmental priming, heartbeat consolidation, and intrinsic
motivation. At the top, A-Evolve~\cite{aevolve} closes the loop with
evolutionary skill refinement, where a Solve--Observe--Evolve--Gate--Reload
cycle optimizes agent skills under monotonic score safety constraints.

\begin{center}
\begin{tikzpicture}[
  layer/.style={draw, minimum width=10cm, minimum height=0.8cm, font=\small},
  >=latex
]
\node[layer, fill=blue!8]  (L5) at (0,3.2) {A-Evolve --- evolution layer};
\node[layer, fill=green!8] (L4) at (0,2.4) {AnimaWorks --- cognitive layer};
\node[layer, fill=yellow!8](L3) at (0,1.6) {AsyncThink --- thinking layer};
\node[layer, fill=orange!8](L2) at (0,0.8) {Ralph\,/\,DeerFlow\,/\,Swarms --- orchestration layer};
\node[layer, fill=red!8]   (L1) at (0,0.0) {Operon --- structural layer};
\end{tikzpicture}
\end{center}

\subsection{ExternalTopology as Universal Intermediate Type}

Every adapter produces an \texttt{ExternalTopology}---a frozen dataclass carrying
the source framework tag, a pattern name, a tuple of agent specifications (plain
dicts with at least \texttt{name} and \texttt{role} keys), directed communication
edges, and arbitrary framework-specific metadata. The analysis pipeline is
source-agnostic: \texttt{analyze\_external\_topology()} consumes any
\texttt{ExternalTopology} and applies the four epistemic theorems
(Theorems~\ref{thm:error-amplification}--\ref{thm:tool-density}) to produce an
\texttt{AdapterResult} containing topology advice, a suggested
\texttt{PatternTemplate}, structural warnings, and a composite risk score. Adding
a new orchestration target therefore requires writing a single parse function---no
changes to the analysis code.

\subsection{Epistemic Theorems as a Structural Linter}

The epistemic theorems from \S\ref{sec:epistemic-topology-section} become a
practical structural linter when applied through the adapter layer. Given an
\texttt{ExternalTopology} with $n$ agents, edge set $E$, and per-agent error
probability $p$, the linter evaluates:

\begin{enumerate}[leftmargin=*]
\item \textbf{Error Amplification} (Theorem~\ref{thm:error-amplification}):
  flags topologies where the aggregate failure probability exceeds a configurable
  tolerance.
\item \textbf{Sequential Penalty} (Theorem~\ref{thm:sequential-penalty}):
  warns when the critical-path length implies excessive latency or compounding
  error risk.
\item \textbf{Parallel Acceleration} (Theorem~\ref{thm:parallel-acceleration}):
  identifies opportunities for parallelism that the current topology leaves
  unexploited.
\item \textbf{Tool Density} (Theorem~\ref{thm:tool-density}):
  detects agents with tool sets exceeding the density threshold, recommending
  specialization.
\end{enumerate}

\noindent For example, analyzing a Swarms \texttt{SequentialWorkflow} with five
agents at $p=0.05$ per agent produces a sequential-penalty warning and suggests
restructuring into a reviewer-gated parallel topology, reducing the expected
error from $1-(1-p)^n \approx 0.23$ to the parallel bound with reviewer
verification.

\subsection{TLA+ Verified Safety Invariants}

Four TLA+ specifications verify safety invariants that are difficult to test
exhaustively at the unit level:

\begin{enumerate}[leftmargin=*]
\item \textbf{TemplateExchangeProtocol}: peer-to-peer template sharing preserves
  provenance, adopted templates carry correct source attribution, and trust scores
  update monotonically in the direction of observed outcomes.
\item \textbf{DevelopmentalGating}: stage transitions are irreversible
  (EMBRYONIC $\to$ JUVENILE $\to$ ADOLESCENT $\to$ MATURE), critical-period
  windows close permanently, and tool acquisition respects the current stage's
  minimum-stage constraint.
\item \textbf{ConvergenceDetection}: the watcher's intervention-count ratio
  converges or triggers a HALT before exceeding the maximum intervention rate,
  preventing unbounded retry loops.
\item \textbf{EvolutionGating}: A-Evolve's evolutionary loop maintains monotonic
  score safety---the deployed skill set's fitness never regresses below the
  pre-evolution baseline.
\end{enumerate}

\noindent Together, these specifications cover the critical safety boundaries of
the convergence stack: trust integrity, developmental ordering, operational
convergence, and evolutionary monotonicity.

\subsection{Co-Design Convergence}

The full adapter stack is formalized as a Zardini co-design
problem~\cite{zardini2023codesign}. Each adapter is a Design Problem (DP)---a
monotone map from a resource poset (available agents, tools, context budget) to a
functionality poset (task coverage, error bounds, latency). Series composition
($\texttt{compose\_series}$) chains adapters where one DP's functionality becomes
the next's resource; parallel composition ($\texttt{compose\_parallel}$) runs
independent DPs on disjoint resource slices. The adaptive assembly loop (run
$\to$ record $\to$ score $\to$ select) is feedback composition: a fixed-point
iteration over the PatternLibrary's scoring function. Convergence is guaranteed
when the scoring function is monotone and the template space is finite---both
hold by construction.

\subsection{Adapter Summary}

Table~\ref{tab:convergence-adapters} summarizes the five adapters, their parse
entry points, and the Operon concepts they map to.

\begin{table}[ht]
\centering
\small
\caption{Convergence adapters: framework, parse function, and Operon mapping.}
\label{tab:convergence-adapters}
\begin{tabular}{@{}p{2.0cm}p{3.6cm}p{7.0cm}@{}}
\toprule
\textbf{Framework} & \textbf{Parse Function} & \textbf{Operon Mapping} \\
\midrule
Swarms \cite{swarms}
  & \texttt{parse\_swarm\_\allowbreak topology()}
  & Workflow patterns $\to$ ExternalTopology, PatternTemplate seeding \\[3pt]
DeerFlow \cite{deerflow}
  & \texttt{parse\_deerflow\_\allowbreak session()}
  & Session configs $\to$ ExternalTopology, bidirectional skill bridge \\[3pt]
AnimaWorks \cite{animaworks}
  & \texttt{parse\_animaworks\_\allowbreak org()}
  & Org hierarchies $\to$ ExternalTopology, memory bridge to BiTemporalMemory \\[3pt]
Ralph \cite{ralph}
  & \texttt{parse\_ralph\_\allowbreak config()}
  & Hat definitions $\to$ ExternalTopology, cognitive-mode--annotated StageSpec \\[3pt]
A-Evolve \cite{aevolve}
  & \texttt{parse\_aevolve\_\allowbreak workspace()}
  & Workspace manifests $\to$ ExternalTopology, evolved skill $\to$ StageSpec \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Current Status}

The convergence package comprises 14 Python modules (including the shared type
definitions and the \texttt{\_\_init\_\_} re-exports) and 216 convergence-specific
tests. The repository includes 98 examples (numbered 1--98), of which examples
86--98 demonstrate convergence
features: adapter analysis (86--88), template exchange and hybrid assembly
(89--91), memory bridging, priming views, heartbeat consolidation, and
asynchronous thinking (92--95), co-design composition (96), and the Ralph and
A-Evolve adapter workflows (97--98). All four TLA+ specifications pass model
checking with TLC.

\subsection{Relationship to the Main Framework}

The convergence layer depends on Operon's internal packages
(\texttt{core}, \texttt{patterns}, \texttt{memory}, \texttt{organelles})
but no core module depends on \texttt{operon\_ai.\allowbreak convergence}.
This one-way dependency ensures that the structural guarantees developed in
the preceding sections remain valid independently of whether any external
adapter is present.
The adapters are consumers of the epistemic analysis API, not extensions
of it.

For the complete treatment---including adapter implementation details, template
exchange protocols, memory bridge semantics, TLA+ specification listings, and all
98 worked examples---see the standalone convergence companion
paper.\footnote{\url{https://coredipper.github.io/operon/convergence/}}
