\begin{abstract}
Reflective-evolution optimizers such as GEPA evolve text components
(prompts, code, configs) by scoring candidates with a user-supplied
evaluator and feeding the results, plus free-form feedback, to a
reflection language model that proposes mutations.  The evaluator
traditionally returns a scalar reward.  We test an alternative in
which the evaluator returns a binary pass/fail verdict paired with
structured obligation text drawn from a formal theorem verifier
(``Operon certificates'').  The conjecture is that, on tasks with
enumerable failure modes, binary-plus-obligations converges in fewer
mutations than a scalar reward of comparable information density.

We run three arms on a synthetic windowed-stability task
(\texttt{behavioral\_stability\_windowed}): \emph{cert-binary} (the
treatment), \emph{scalar-reward} (a graded reward that peaks when the
theorem holds), and \emph{scalar-with-evidence} (an active control
carrying the same per-window violation text as the cert arm but
without the theorem framing).  Ten seeds per arm, 50-iteration
budget, Gemma~4 as the reflection LM.  All three arms converged on
100\% of runs (Wilson 95\% CI $[0.722, 1.000]$ per arm).  Mean
iterations-to-convergence were 1.0 for cert-binary, 2.0 for
scalar, and 1.0 for scalar-evidence.  Against the strongest
baseline (scalar-evidence), cert-binary is statistically
indistinguishable (Mann-Whitney $U=50$, $p=1.000$, rank-biserial
$0.00$); against the weaker baseline (scalar), cert-binary is
faster ($U=35$, $p=0.039$, rank-biserial $0.30$).  The predeclared
gate (ratio $\le 0.75$ at $p<0.05$ vs the stronger baseline) is
not cleared, so the result is a \emph{null} for the binarization
hypothesis: the per-window obligation evidence carries the signal,
the binary/theorem framing does not add on top.

The ablation on the obligation formatter isolates whether
the signal comes from binarization or from obligation content.  The
work is an engineering test of a theoretical claim (Theorem~3 of the
companion integration memo) rather than a new algorithm; the
deliverable is an honest measurement, and negative results are
reported as such.
\end{abstract}
