\section{Limitations}\label{sec:limits}

\paragraph{Single theorem.}  Only
\texttt{behavioral\_stability\_windowed} is tested.  The conjecture
(memo Theorem~3) applies to theorems with \emph{enumerable} failure
modes; generalization to, e.g., \texttt{behavioral\_quality}
(continuous-score theorems) or \texttt{no\_false\_activation}
(structural theorems) is untested.  If the primary result is
positive, a follow-up sweep across at least
\texttt{behavioral\_quality} and \texttt{behavioral\_no\_anomaly}
is the natural next experiment.

\paragraph{Single LM.}  The reflection model is fixed at Gemma~4.
Frontier models may parse structured obligation text differently.
A cross-LM re-run on the treatment and strongest baseline is
declared as follow-up.

\paragraph{Single mutable component, by design.}  Our candidate has
two keys --- an immutable prose scratch-space
(\texttt{policy\_prompt}) and a mutable numeric knob
(\texttt{policy\_throttle}) --- but a custom
\texttt{module\_selector} (\texttt{\_ThrottleOnlySelector}) pins
GEPA's component selection to the throttle, so only one degree of
freedom is searched.  This matches the task's one-knob structure
but means GEPA's component-selection machinery is exercised
trivially.  Multi-component candidates --- where different
components fail different obligations and GEPA must choose which to
mutate --- may change the picture, both for the adapter's reflection
formatting (multiple components imply multiple per-component
templates) and for the cert-binary signal (a binary verdict on a
multi-component candidate fans out to which component to blame).

\paragraph{Synthetic task.}  The harness is deliberately a
one-knob task to isolate the evaluator-to-optimizer pipeline.  A
strong result here suggests but does not establish that real
agent tasks will behave the same way; realistic tasks introduce
retrieval noise, API flakiness, and agent-complexity variance
that may interact with either arm.

\paragraph{GEPA version coupling.}  Results are reported for GEPA
\texttt{0.1.1}.  GEPA's adapter protocol may evolve; the
\texttt{propose\_new\_texts} attribute, for instance, is a required
member even when unused.  Future GEPA versions may change the
conditions under which certificate-based adapters are picked up.

\paragraph{Mock-smoke and real-run separation.}  The experiment
driver supports a mock reflection LM for CI smoke testing.  Mocked
runs are flagged in the output JSON (\texttt{mock\_reflection\_lm:
true}) and filtered by the analysis script by default.  Measured
results in Section~\ref{sec:results} use real Gemma~4 only.

\paragraph{No compute-adjusted comparison.}  The arms use the same
budget in GEPA iterations, but the reflection-LM token cost differs
by arm (the cert-binary feedback is longer than the scalar arm's
single-line score).  A compute-adjusted re-run, matching total
reflection-LM token budget rather than iteration count, is a
follow-up.
