\section{Conclusion}\label{sec:conclusion}

We tested Theorem~3 of Operon's external-frameworks integration
memo --- the conjecture that certificate-based evaluators converge
faster than scalar-reward evaluators on tasks with enumerable
failure modes --- by running GEPA with three evaluator shapes on a
synthetic windowed-stability task.  The cert-binary arm replaces
the scalar reward with a pass/fail verdict plus structured
obligation text drawn from an Operon theorem verifier.  The
scalar-reward and scalar-with-evidence arms form a minimal pair
that isolates binarization from obligation-text content.

\textbf{Verdict: refine.}  All three arms reach 100\% convergence
within budget; mean iterations-to-convergence are 1.0
(cert-binary), 2.0 (scalar), and 1.0 (scalar-evidence).  Against
the stronger baseline (scalar-evidence) cert-binary is statistically
indistinguishable (Mann-Whitney $U=50$, $p=1.000$, rank-biserial
$0.00$), so the predeclared $\le 0.75$-ratio gate is not cleared
and Theorem~3 stays a conjecture rather than an accepted theorem
for this task.  The formatter ablation on seeds 0--4 ties at 1.0
mean iterations for both the default and the minimal formatter,
localising the active ingredient to ``\emph{some} structured
failure information vs none'' --- neither binarization, theorem
framing, nor the numeric content of the obligation evidence adds
on top.  The integration memo is updated accordingly: certificates
remain a clean evaluator interface, but on this task they bring no
convergence-speed advantage over a graded scalar reward paired with
the same evidence text.

\subsection{Follow-ups unblocked}\label{sec:followups}

The list below is ordered by dependency: items 1--2 reuse this
paper's protocol verbatim and are the cheapest to launch; items
3--5 are independent of each other and have no dependency on items
1--2.  Each carries a target metric and a binary done criterion so
the work can be picked up out of order without re-deriving scope.

\begin{enumerate}[leftmargin=*,itemsep=2pt,topsep=2pt]
\item \textbf{Theorem-shape sweep.}  Rerun the protocol on
\texttt{behavioral\_quality} (continuous) and
\texttt{behavioral\_no\_anomaly} (structural).
\emph{Prerequisites:} none --- the harness, adapter shapes, and
analysis pipeline ship in this paper.
\emph{Target:} the predeclared $\le 0.75$-ratio gate on each
theorem shape.
\emph{Done:} paper-6-shaped writeup per theorem, with a
per-shape verdict against the gate.
\item \textbf{Cross-LM robustness.}  Rerun cert-binary and the
strongest baseline (scalar-with-evidence) with at least one
frontier LM in addition to Gemma~4.
\emph{Prerequisites:} budget for non-local LM tokens; otherwise
identical infrastructure.
\emph{Target:} same $\le 0.75$-ratio gate, evaluated independently
per LM; flag any qualitative arm-ordering flip.
\emph{Done:} an LM~$\times$~arm table reported alongside
Table~\ref{tab:paper6-iters}.
\item \textbf{Operon--autocontext adapter.}  Wire Operon certificates
into autocontext's staged-validation loop~\cite{autocontext2025},
giving a second external framework that consumes certificates as
evaluator output.  A plan for this is logged separately; this is a
paper-7 candidate.
\emph{Prerequisites:} autocontext's staged-validation hook
surface (independent of items 1--2).
\emph{Target:} an in-tree
\texttt{AutocontextCertificateAdapter} that mirrors
\texttt{GascityCertificateAdapter}'s shape, with one dogfood
test that consumes an Operon theorem through the autocontext
boundary.
\emph{Done:} adapter merged in
\texttt{operon\_ai/convergence/} with the dogfood test green.
\item \textbf{Lightweight DSPy reproducibility theorem.}  Ship
\texttt{dspy\_compile\_pinned\_inputs} as the cheap provenance
variant of memo Theorem~2 --- pinned-input hashes recorded at
compile time, no recompile.
\emph{Prerequisites:} stable
\texttt{Certificate.from\_dspy\_compile} entry point (already
shipped 2026-05-01 per memo §2).
\emph{Target:} register the verify function on the theorem
registry and exercise it from a DSPy compile.
\emph{Done:} theorem in
\texttt{operon\_ai/core/certificate.py}, unit test
exercising the round-trip, memo §2 \emph{Status} line updated.
\item \textbf{\texttt{gepa\_candidate\_improvement} theorem.}
Certify that an evolved candidate Pareto-dominates its parent on
the validation set; close GEPA's outer loop with a certificate.
\emph{Prerequisites:} GEPA exposes per-iteration parent +
child candidate pairs (already available via callbacks per this
paper's \texttt{\_IterationRecorder}).
\emph{Target:} a verify function whose evidence is the
$(\text{parent valset scores}, \text{child valset scores})$
pair; cert-binary arm on a held-out task.
\emph{Done:} theorem registered, a regression
test that constructs a known-dominating child and a known-tied
child and verifies both verdicts.
\end{enumerate}

\subsection{Artifacts}

\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
\item Experiment driver:
\texttt{eval/convergence/theorem\_6\_experiment.py}
\item Ablation driver:
\texttt{eval/convergence/theorem\_6\_ablation.py}
\item Harness:
\texttt{eval/convergence/synthetic\_signal\_harness.py}
\item Baseline adapters:
\texttt{scalar\_reward\_adapter.py},
\texttt{scalar\_with\_evidence\_adapter.py}
\item Analysis:
\texttt{eval/convergence/analysis\_theorem\_6.py}
\item Per-arm trajectories:
\texttt{eval/results/theorem\_6/\{arm\}/seed\_*.json}
\item Ablation trajectories:
\texttt{eval/results/theorem\_6/ablation\_minimal/seed\_*.json}
\item Summaries:
\texttt{eval/results/theorem\_6/summary.json},
\texttt{eval/results/theorem\_6/ablation\_summary.json}
\item Unit tests:
\texttt{tests/unit/convergence/test\_paper6\_*.py}
\end{itemize}

\paragraph{Reproducibility.}
\begin{verbatim}
pip install gepa matplotlib scipy
ollama pull gemma4  # if not already local
python -m eval.convergence.theorem_6_experiment --sweep
python -m eval.convergence.theorem_6_ablation
python -m eval.convergence.analysis_theorem_6
cd article/paper6 && tectonic main.tex
\end{verbatim}
